---
title: "VC55 Weather Case Study"
author: Paul J. Palmer
date: March 22, 2005
output: beamer_presentation
---
```{r setDefaults,  include=TRUE, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE, results='asis'}
# Set code chunk options for all chunks
# These can be overridden at the chunk level, but setting global options 
# ensures consistency of chunk behaviour.
# To print a version of this document without code set echo = FALSE. 
#Include false will ignore the code chunk completely!
knitr::opts_chunk$set(include = TRUE, 
                      echo = TRUE, 
                      warning=FALSE, 
                      message=FALSE, 
                      error=FALSE, 
                     # fig.width = 3,
                    #  fig.height = 2,
                     # fig.width = 3,
                    out.width = '70%',
                      fig.align = "center",
                      results='asis')



```


## Introduction

   - The purpose of this vignette is a practical demonstration of reusable templates based upon the novel concept of data state. 
   - Pages from this report are used to create Figures describing a case study for the paper: *Achieving Analytical Fluency With Complex Data*.
   -  Real world long term weather data are used as the source. 
   - **It is not the intention to analyse climate change, but the trends uncovered are striking, even using this single public domain data source**.

## Load Libraries

Load the Tidyverse libraries and other helpers before the analysis starts.


```{r }
# Load libraries
library(tidyverse) # Load all the Tidyverse packages in one go.
library(kableExtra) # Enable advanced table styling
library(qqplotr) # Enable QQ plots for statistical analysis
library(readr) # Improved reading of text files
library(timeDate) # Helper for time date manipulation
library(rprojroot) # Useful utilities
library(fs) # System independent file paths

```



